Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
Small changes I had noted

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation (\cite{witten2016data}).

@@

These imputation algorithms are used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset
(\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be considered as another avenue of assessing
imputation confidence.

>>>>

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained-equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation (\cite{witten2016data}).

@@

These imputation algorithms are used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset
(\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be another avenue of assessing
imputation confidence.

### 2017-03-01 paper/experiments.tex
Small changes I had noted

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0.0$),
\ProjectName{} optimizing for performance ($\alpha=1.0$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

@@

Each query was separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach is low in all cases---between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent---indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

We also calculate the number of tuples used to produce each aggregate output.
The Count columns in \Cref{table:smape} show the percentage difference between the number of tuples in the aggregate for $\alpha = 1$ and $\alpha = 0$ and running the query on the imputed base table.
This shows that when optimizing for performance, not quality, many tuples are dropped.
Even in cases where the SMAPE reduction between $\alpha = 1$ and $\alpha = 0$ is small (\Cref{q2}, \Cref{q6}) the tuple count is significantly different.
In these cases, the aggregate value is not significantly impacted by the missing data.
However, if the missing data is biased then the aggregate will have a significant error.
This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
By using \ProjectName{} this cost can be lowered significantly.

>>>>

Each query was separately run 200 times in both settings and then

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

@@

For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach is low---between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent---indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

We also calculate the number of tuples used to produce each aggregate output.
The Count columns in \Cref{table:smape} show the percentage difference between the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ and running the query on the imputed base table.
This shows that when optimizing for performance, not quality, many tuples are dropped.
Even in cases where the SMAPE reduction between $\alpha = 1$ and $\alpha = 0$ is small (\Cref{q2}, \Cref{q6}) the tuple count is significantly different.
In these cases, the aggregate value is not significantly impacted by the missing data. In particular, if values are missing uniformly at random,
the aggregate should not be affected.
However, if the missing data is biased then the aggregate will have a significant error.
This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
By using \ProjectName{} this cost can be lowered significantly.
